### 2018-06-12 express2018/formal.tex
text

In the second half of this paper, we present a formalization of CCS
including the new concepts and theorems proposed in the first half of
this paper. The main purpose is to convince the readers that, there's no flaw in the informal proofs. The
formalization work is based on a formalization of CCS in HOL theorem
prover (HOL4) \cite{slind2008brief}, described in second author's MSc thesis
\cite{Tian:2017wrba} and now lives in the official examples folder of HOL4 source
code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}.
The work is essentially a porting and extension of the early CCS
formalization by Monica Nesi \cite{Nesi} in HOL88.

Higher Order Logic (or ``HOL Logic'') \cite{hollogic} stands for simple-typed $\lambda$-calculus plus Hibert
choice operator, axiom of infinity, and rank-1 polymorphim (type
variables). HOL4 implemented this logic without any extension, while
some other theorem provers in HOL family (e.g. Isabelle/HOL) has
various extensions. The HOL Logic is considerability simpler than logic
foundations in most other theorem provers, e.g. Coq. As a result,
formal theories in HOL can be easily migrated (sometimes even
automatically) into other theorem provers.

We will not repeat all definitions and theorems of CCS again by their
formalized versions. Instead, we just focus on several highlights in
this work, i.e.
\begin{enumerate}
\item The formalization of semantics context in $\lambda$-expressions and the theory of
  (pre)congruence for CCS;
\item The definition and uses of trace in the proof of unique solution of
  contractions theorem;
\item The proof of the hard version of ``coarsest congruence
  contained in $\approx$'' theorem.
\end{enumerate}

\subsection{Simplifications and Limitations}

We have only formally verified simplified forms of the unique
solution theorems introduced in the first part. This is to not only
save the formalization efforts, but also to make us focusing on the most
important part of the proof. The most important simplification is
that, we have only proved all unique solution theorems with only single-variable equations
(contractions).

But such a simplification doesn't hurt the rigorousness of our entire
work, as once the single-equation version of the theorem is
proven, the general case is just a routine
adaptation for paper proofs.
It's not easy to formalize multi-variable CCS equations with
introducing a large number of definitions and intermediate
results. It's easy, however, to directly use the semantics context to
represent single-variable equations without the need of formalizing
the concept of ``equation'' at all.

There're two major limitations in this work. One is due to HOL, we only support Finitary CCS, i.e.
no infinite sum nor infinite parallel composition.
The other limitation is, equation variables (or holes in semantics contexts)
cannot appear inside recursion operator closure. However, these limitations do not
cause any difference in the proof of main results represented in this paper.

>>>>

We highlight here a formalization of CCS
in the HOL theorem
prover (HOL4) \cite{slind2008brief},
including the new concepts and theorems proposed in the first half of
this paper.
%  The main purpose is to convince the readers that, there's no flaw
%  in the informal proofs. 
The whole formalisation can be found 
in \cite{Tian:2017wrba}  or in 
 in the official example folder of HOL4 source
code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}.
\finish{i would say here how big is the formalisation, ie how many
  lines of code} 
%  This
%  work is based on a formalization of CCS in HOL theorem
% prover (HOL4) \cite{slind2008brief}, described in second author's MSc thesis
% \cite{Tian:2017wrba} and now lives in the official examples folder of HOL4 source
% code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}.

Higher Order Logic (or ``HOL Logic'') \cite{hollogic} stands for simple-typed $\lambda$-calculus plus Hibert
choice operator, axiom of infinity, and rank-1 polymorphim (type
variables). HOL4 implemented this logic without any extension, in
contrast
with
some other theorem provers of the  HOL family (e.g. Isabelle/HOL). 
\finish{not having extensions above: sounds like a defect; what is the
advantage? the portability explained in the next sentence?} 
The HOL Logic has considerability simpler logic
foundations than most other theorem provers, e.g. Coq. As a result,
formal theories in HOL can be easily migrated (sometimes even
automatically) into other theorem provers.

\finish{below: to be reformulated} 
We will not repeat all definitions and theorems of CCS again by their
formalized versions. Instead, we just focus on several highlights in
this work, i.e.
\begin{enumerate}
\item The formalization of semantics context in $\lambda$-expressions and the theory of
  (pre)congruence for CCS;
\item The definition and uses of trace in the proof of unique solution of
  contractions theorem;
\item The proof of the hard version of ``coarsest congruence
  contained in $\approx$'' theorem.
\end{enumerate}

In contrast with the theorems in the first part of the paper, in the
formalisation we consider only single equations/contractions. 
This considerably simplifies the  reading of the HOL scripts. 
The extension of the work to multiple equations/contractions is
discussed in Section~\ref{s:}.

% \subsection{Simplifications and Limitations}

% We have only formally verified simplified forms of the unique
% solution theorems introduced in the first part. This is to not only
% save the formalization efforts, but also to make us focusing on the most
% important part of the proof. The most important simplification is
% that, we have only proved all unique solution theorems with only single-variable equations
% (contractions).

% But such a simplification doesn't hurt the rigorousness of our entire
% work, as once the single-equation version of the theorem is
% proven, the general case is just a routine
% adaptation for paper proofs.
% It's not easy to formalize multi-variable CCS equations with
% introducing a large number of definitions and intermediate
% results. It's easy, however, to directly use the semantics context to
% represent single-variable equations without the need of formalizing
% the concept of ``equation'' at all.

% There're two major limitations in this work. One is due to HOL, we only support Finitary CCS, i.e.
% no infinite sum nor infinite parallel composition.
% The other limitation is, equation variables (or holes in semantics contexts)
% cannot appear inside recursion operator closure. However, these limitations do not
% cause any difference in the proof of main results represented in this paper.

### 2018-06-12 express2018/sos.tex
text

In the formalization of CCS, it's better to have two distinct
types for actions: the type \HOLinline{\ensuremath{\beta} \HOLTyOp{Label}} ($\beta$ is a type variable.) for visible actions, divided by input
and output actions:
\begin{lstlisting}
val _ = Datatype `Label = name 'b | coname 'b`;
\end{lstlisting}
and the type \HOLinline{\ensuremath{\beta} \HOLTyOp{Action}} as the
union of all visible and invisible actions. The cardinality of all
available types depends on the choice of type variable $\beta$.

The core datatype ``\HOLinline{(\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS}}'' is then defined as an inductive
datatype in HOL, based on its Datatype package:
\begin{lstlisting}
val _ = Datatype `CCS = nil
		      | var 'a
		      | prefix ('b Action) CCS
		      | sum CCS CCS
		      | par CCS CCS
		      | restr (('b Label) set) CCS
		      | relab CCS ('b Relabeling)
		      | rec 'a CCS`;
\end{lstlisting}

We have added some minimal grammar support using HOL's powerful pretty printer, to represent CCS
processes in more readable forms. Table \ref{tab:ccsoperator} has listed the notation of typical CCS processes and
major operators supported by above definition. (Notice the use of
recursition operator for representing process constants)
\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Operator} & \textbf{Notation} & \textbf{HOL term} & \textbf{HOL
  (alternative)}\\
\hline
nil & $\textbf{0}$ &  \HOLinline{\HOLConst{nil}} & \HOLinline{\HOLConst{nil}} \\
Prefix & $a.b.0$ & \texttt{prefix a (prefix b nil)} & \HOLinline{\HOLFreeVar{a}\HOLSymConst{..}\HOLFreeVar{b}\HOLSymConst{..}\HOLConst{nil}} \\
Sum & $p + q$ & \texttt{sum p q} & \HOLinline{\HOLFreeVar{p} \HOLSymConst{+} \HOLFreeVar{q}} \\
Parallel & $p \,\mid\, q$ & \texttt{par p q} & \HOLinline{\HOLFreeVar{p} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{q}} \\
Restriction & $(\nu\;L)\;p$ & \texttt{restr L p} & \HOLinline{\HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} \HOLFreeVar{p}} \\
Constant & $A=a.A$ & \texttt{rec A (prefix a (var A))} & \HOLinline{\HOLConst{rec} \HOLFreeVar{A} (\HOLFreeVar{a}\HOLSymConst{..}\HOLConst{var} \HOLFreeVar{A})} \\
\hline
\end{tabular}
\end{center}
\vspace{-1em}
   \caption{Syntax of CCS operators}
   \label{tab:ccsoperator}
\end{table}

The transition semantics of CCS processes were defined by the following
Structural Operational Semantics (SOS for short) rules in
Fig.\;\ref{f:LTSCCS}, except that the transition rule for constants
has been changed to recursion operator:
\HOLTokenTurnstile{} \HOLConst{CCS_Subst} \HOLFreeVar{E} (\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}) \HOLFreeVar{X} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}
which says, if we substitute all appearences of constant $X$ in $E$ to
$(\mathbf{rec}\; X\; E)$ and the resulting process has a transition to $E_1$
with action $u$, then $(\mathbf{rec}\; X\; E)$ has the same
transition. Other SOS rules have the following forms:

@@

@@

Above rules work as \emph{axioms} in the theory of CCS, as they cannot be
proved by more primitive theorems or axioms. From the view of theorem prover, these
SOS rules are nothing but an \emph{inductive 
  definition} on 3-ary relation \HOLinline{\HOLConst{TRANS}} of type ``\HOLinline{(\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{\beta} \HOLTyOp{Action} \HOLTokenTransEnd (\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS} \HOLTokenTransEnd \HOLTyOp{bool}}'', generated by HOL4's function
\texttt{Hol_reln}.\footnote{On the other side, above literal SOS rules are not
the \emph{only} thing we got from \texttt{Hol_reln}, which also
returns the fix-point theorem and induction principle for the
relation, otherwise things like ``$\textbf{0}$ has no transition''
cannot be proved at all.} The ultimate result here (before
bisimilution gets involved), is an ML function taking a CCS process
and returning a theorem indicating all its direct transitions.

>>>>

In the formalization of CCS, 
the type \HOLinline{\ensuremath{\beta} \HOLTyOp{Label}} ($\beta$ is a
type variable) accounts for visible actions, divided into input
and output actions:
\begin{lstlisting}
val _ = Datatype `Label = name 'b | coname 'b`;
\end{lstlisting}
Then
the type \HOLinline{\ensuremath{\beta} \HOLTyOp{Action}} is the
union of all visible and invisible actions. The cardinality of
\HOLinline{\ensuremath{\beta} \HOLTyOp{Action}} (and therefore of all
CCS types built on top of it) 
 depends on the choice of initial  type $\beta$.
\finish{i have reformulated the sentence above; check it}

The core datatype ``\HOLinline{(\ensuremath{\alpha},
  \ensuremath{\beta}) \HOLTyOp{CCS}}'', accounting for the CCS syntax,  is then defined as an inductive
datatype in HOL, based on its Datatype package:
\finish{above: check the change i made. Also, explain what is $\alpha
  $ (i suppose the set of  variables for recursion)} 
\begin{lstlisting}
val _ = Datatype `CCS = nil
		      | var 'a
		      | prefix ('b Action) CCS
		      | sum CCS CCS
		      | par CCS CCS
		      | restr (('b Label) set) CCS
		      | relab CCS ('b Relabeling)
		      | rec 'a CCS`;
\end{lstlisting}

We have added some  grammar support using HOL's powerful pretty printer, to represent CCS
processes in more readable forms. Table \ref{tab:ccsoperator} has listed the notation of typical CCS processes and
major operators supported by above definition.

\finish{explain what is ``HOL   (alternative)''. 
Also: the table gives the impression that all recursion are guarded,
ie, the body of the recursion begins with a prefix. I would instead
put in the table the plain recursion rec A.P. Also, in the row for
prefix, i would write plain prefix u.P}
\finish{explain why restriction takes a set of names, rather than a
  single name as in the  CCS syntax}

%  (Notice the use of
% recursition operator for representing process constants)
\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Operator} & \textbf{CCS Notation} & \textbf{HOL term} & \textbf{HOL
  (alternative)}\\
\hline
nil & $\textbf{0}$ &  \HOLinline{\HOLConst{nil}} & \HOLinline{\HOLConst{nil}} \\
Prefix & $a.b.0$ & \texttt{prefix a (prefix b nil)} & \HOLinline{\HOLFreeVar{a}\HOLSymConst{..}\HOLFreeVar{b}\HOLSymConst{..}\HOLConst{nil}} \\
Sum & $p + q$ & \texttt{sum p q} & \HOLinline{\HOLFreeVar{p} \HOLSymConst{+} \HOLFreeVar{q}} \\
Parallel & $p \,\mid\, q$ & \texttt{par p q} & \HOLinline{\HOLFreeVar{p} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{q}} \\
Restriction & $(\nu\;L)\;p$ & \texttt{restr L p} & \HOLinline{\HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} \HOLFreeVar{p}} \\
Recursion & $A=a.A$ & \texttt{rec A (prefix a (var A))} & \HOLinline{\HOLConst{rec} \HOLFreeVar{A} (\HOLFreeVar{a}\HOLSymConst{..}\HOLConst{var} \HOLFreeVar{A})} \\
\hline
\end{tabular}
\end{center}
\vspace{-1em}
   \caption{Syntax of CCS operators}
   \label{tab:ccsoperator}
\end{table}

The transition semantics of CCS processes  follows
the Structural Operational Semantics (SOS) rules in
Figure~\ref{f:LTSCCS}:

@@

\HOLTokenTurnstile{} \HOLConst{CCS_Subst} \HOLFreeVar{E} (\HOLConst{rec} \HOLFreeVar{A} \HOLFreeVar{E}) \HOLFreeVar{A} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{rec} \HOLFreeVar{A} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}
\texttt{[RECURSION]}

@@

The rule 
\texttt{[RECURSION]
 says that if we substitute all appearences of variable $A$ in $E$ to
$(\mathbf{rec}\; A\; E)$ and the resulting process has a transition to $E_1$
with action $u$, then $(\mathbf{rec}\; A\; E)$ has the same
transition.

% Above rules work as \emph{axioms} in the theory of CCS, as they cannot be
% proved by more primitive theorems or axioms.

From the HOL  viewpoint, these
SOS rules are \emph{inductive 
  definitions} on the 3-ary relation \HOLinline{\HOLConst{TRANS}} of type ``\HOLinline{(\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{\beta} \HOLTyOp{Action} \HOLTokenTransEnd (\ensuremath{\alpha}, \ensuremath{\beta}) \HOLTyOp{CCS} \HOLTokenTransEnd \HOLTyOp{bool}}'', generated by HOL4's function
\texttt{Hol_reln}.\footnote{\texttt{Hol_reln} also
gives us a fix-point theorem and induction principle for the
relation, otherwise statements like ``$\textbf{0}$ has no transition''
could not be proved.}

\finish{the text below is important but there are various things that
  are not clear: where does ML come from (we have not talked about ML
  before and for a non-expert in HOL this is a mystery)? did you have
  to define it? was it difficult or is there some HOL facility that helps? }

The ultimate result here (before
bisimilution gets involved), is an ML function
\finish{is this the function \texttt{CCS\_TRANS\_CONV}}  mentioned
later below? }
  taking a CCS process
and returning a theorem indicating all its direct transitions.
